Missing values

### 2021-10-27 lit_review.Rmd
updating gitignore

---
title: "Literature review: Inequality and the biosphere"
author: "Juan Rocha"
date: "May 2021"
output:
  html_notebook:
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
    selfcontained: true
---

@@

library(tidyverse)
library(tidytext)
library(tm)
library (topicmodels)
library(lda)
library(naniar)
library(plotly)
library(patchwork)

theme_set(theme_light(10))
```

Below I read the data and organize the data by most citations with the mock data (N=2000) that Emmy provided:
```{r}
dat <- read_csv(file = "data/April27_1-2000.csv") %>% 
    janitor::clean_names()
dat %>% 
    select(title, authors, cited_by, year, author_keywords, abbreviated_source_title) %>% 
    arrange(desc(cited_by))
```

@@

The topic model below is a quick and rough analysis. A proper one would need a sensitivity analysis over a number of choices that I'm making here arbirtrarily such as the algorithm to use, number of topics to fit, etc. Works for brainstorming, do not trust the end result.

```{r}
## extra words
too_words <- tibble(
    word = c("paper", "study", "aim", "aims", "objective", "purpose")
)

dtm <- dat %>% 
    select(abstract, title, year) %>%
    filter(!is.na(abstract)) %>%
    unique() %>% # 2193
    unnest_tokens(word, abstract) %>% 
    filter(!is.na(word)) %>% ## Books don't have abstract so they get dropped (n = 8)
    anti_join(stop_words) %>% 
    anti_join(too_words) %>%
    filter(!str_detect(word, "[:digit:]")) %>%
    group_by(title) %>%
    count(word, sort = TRUE) %>%
    cast_dtm(document = title, term = word, value = n)

dtm

@@

Now we have a document term matrix with `r dim(dtm)[1]` documents (abstracts) and `r dim(dtm)[2]` words, after cleaning for common words, stop words and numbers. I pre-computed the topic models outside this notebook because it takes some time and memory, so not suitable for computing at runing time. Here are some visualizatins of the preliminary results:

```{r}
load("data/topic_models_gibbs.RData")
k <- c(5,10,25,50,100)
df_topic_number <- tibble(
    topic_number = k,
    entropy = map_dbl (topicNumber.TM, function (x)
        mean(apply(posterior(x)$topics, 1, function (z) - sum(z * log(z)))) # maximize Entropy
    ),
    alpha = map_dbl(topicNumber.TM, slot, "alpha"),
    log_lik = map_dbl(topicNumber.TM, logLik) #,  #maximize loglik
    #perplexity = map_dbl(topicNumber.TM, perplexity) #minimize perplexity
)

df_stats <- tibble(
    model = names(lapply(tset.TM, logLik)),
    loglik = as.numeric(lapply(tset.TM, logLik)), #maximize loglik
    entropy = lapply (tset.TM, function (x) 
        mean(apply(posterior(x)$topics,
                   1, function (z) - sum(z * log(z))))) %>% as.numeric()#maximize ENTROPY

)

perp <-  lapply(tset.TM[c(1,2,4)], perplexity)
perp$Gibbs <- NA
# pretty names:
df_stats$model <- c("VEM alpha", "VEM fixed", "Gibbs", "CTM")

g1 <- df_stats %>%
    add_column(perplexity = as.numeric(perp[c(1,2,4,3)])) %>% #minimize perplexity
    pivot_longer(cols = 2:4, names_to = "measure", values_to = "value") %>%
    ggplot(aes(model, value)) + 
    geom_col() + 
    # scale_y_continuous(labels = scales::label_scientific) +
    facet_wrap(.~measure, scales = "free_y") +
    labs(x = "Algorithm", y = "Value", tag = "A") +
    theme_light(base_size = 8) + 
    theme(axis.text.x = element_text(size = 5))

g2 <- df_topic_number %>%
    # mutate(alpha_log = log10(alpha)) %>%
    pivot_longer(cols = 2:last_col(), names_to = "measure", values_to = "value") %>%
    # filter(measure != "alpha") %>%
    ggplot(aes(as.factor(topic_number), value)) +
    geom_col() + 
    # scale_y_continuous(labels = scales::label_scientific) +
    labs(x = "Number of topics", y = "Value", tag = "B") +
    facet_wrap(.~measure, scales = "free", ncol = 4, nrow = 1) +
    theme_light(base_size = 8)

g1/g2
```

The best algorithm should maximize entropy and the log-likelihood while reducing perplexity. The latter cannot be calculated for Gibbs sampling, but with the small dataset we have, it seems to be the best performing algorithm. The optimal number of topics is something between 25 and 50, but not higher than 100, where the log-likelihood starts to decay again. For now I will be using 25 topics and Gibbs sampling for the rest of the analysis.

The heatmap below shows how a few documents (on the rows) have a high content of certain topics (on columns).

```{r}
gplots::heatmap.2(
    posterior(topicNumber.TM[[3]])$topics, 
    key=T, trace="none", margins=c(5,5), 
    cexRow=0.7, cexCol=0.7, col="topo.colors", keysize=1, key.par = list(cex = 0.5),
    labRow = FALSE, key.title = NA, key.xlab = "probability", key.ylab = "count" )
```

And below are the topics. A topic is the (posterior) probability distributions of words being likely to be representing together certain hidden structures (topics) in the text. All words in the vocabulary (N = `r dim(dtm)[2]` words) have a probability of explaining a topic, here I'm only showing the top 10 words per topic:

```{r fig.width=8, fig.height=6}
df_topics25 <- tidy(topicNumber.TM[[3]], matrix = "beta")

g_25 <- df_topics25 %>%
    group_by(topic) %>%
    top_n(10, beta) %>% 
    ungroup() %>%
    arrange(topic, - beta) %>%
    mutate(term = reorder_within(term, beta, topic)) %>%
    ggplot(aes(term, beta)) +
    geom_col(aes(fill = as.factor(topic)),show.legend = FALSE) +
    coord_flip() + 
    scale_x_reordered() +
    labs(tag = "D", y = "Probability of word explaining the topic", x = "Word ranking") +
    facet_wrap(.~ topic, scales = "free_y", ncol = 5) +
    theme_light(base_size = 7) + 
    theme(axis.text.x = element_text(size = 5))

g_25
```

Now let's see how topics change over time:

```{r}
df_documents <- tidy(topicNumber.TM[[3]], matrix = "gamma")

papers_yr <- dat %>%
    #filter(year > 2000, year < 2021) %>%
    group_by(year) %>%
    tally() %>% rename(total_papers = n)

g_stack <-  df_documents %>% 
    left_join(
        dat %>% 
            select(abstract, title, year) %>%
            filter(!is.na(abstract)) %>%
            unique(),
        by = c("document" = "title") ) %>%
    mutate(year = as.numeric(year)) %>% 
    filter(year > 2000, year <2021) %>%
    group_by(topic, year) %>% 
    summarize(gamma_yr_tp = sum(gamma), .groups = "drop") %>% 
    group_by(year) %>%
    mutate(total = sum(gamma_yr_tp)) %>%
    ungroup() %>% 
    mutate(proportion = gamma_yr_tp / total,
           topic = as.factor(topic)) %>%
    ggplot(aes(x = year, y = proportion, group = topic)) +
    geom_area(aes(fill = topic, color = topic), 
              position = "stack", show.legend = FALSE, alpha = 0.5, size = 0.25) +
    labs(y = "Relative proportion of topics", x = "Year") +
    theme_classic(base_size = 10)

g_stack
```

If you're interested in learning more about the method or applications, you can learn more [here](https://arxiv.org/abs/2012.14312) or the references therein.

@@

@@

```{r fig.width=8}
library(network)
authors <- dat %>% 
    select(authors, doi) %>% 
    mutate(authors = str_split(authors, pattern = ", ")) %>% 
    unnest(cols = authors) %>% 
    mutate(presence = 1)

mat <- authors %>% unique() %>% 
    pivot_wider(names_from = authors, values_from = presence, values_fill = 0) %>% 
    select(-doi) %>%
    as.matrix()

mat <- t(mat) %*% (mat)

df_authors <- as_tibble(mat, rownames = "author")

df_authors <- df_authors %>% 
    pivot_longer(2:last_col(), names_to = "co_author", values_to = "n_papers") %>% 
    filter(author != co_author, n_papers > 0)

net <- df_authors %>% 
    # filtering reduces from 232k links and >4.5k authors to 708 links shared by 233 authors
    filter(n_papers > 1) %>% 
    network(., directed = FALSE, matrix.type = "edgelist")

networkD3::simpleNetwork(
    Data= df_authors %>% filter(n_papers >1) ,
    Source = "author", Target = "co_author", 
    zoom = TRUE, linkColour = "grey", nodeColour = "blue"
)

## free memory
rm(mat, authors)

>>>>

---
title: 'Literature review: Inequality and the biosphere'
author: "Juan Rocha"
date: "September 2021"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float:
      collapsed: true
      smooth_control: true
    toc_depth: 3
    fig_caption: true
    fig_width: 8
    fig_height: 5
    highlight: tango
    self_contained: true
    theme: "paper"

  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    theme:
      bootswatch: cosmo
      code_font:
        google: Fira Code
---

@@

library(tidyverse)
library(tidytext)
library(tm)
library(topicmodels)
library(lda)
library(naniar)
library(plotly)
library(patchwork)

theme_set(theme_light())
```

Below I read the data and organize the data by most citations with the data that Emmy provided (N=86430 papers):
```{r message = FALSE, warning=FALSE}

fls <- list.files(path = "data/lit_review/")

dat_xls <- paste0("data/lit_review/", str_subset(fls, ".xlsx")) %>% 
    map(readxl::read_excel) %>% 
    map(janitor::clean_names) %>% 
    map(function(x) x %>% select(abstract, title, year, doi, cited_by) %>% as_tibble()) %>% 
    map(function(x) x %>% mutate(across(.cols = everything(), .fns = as.character))) %>% 
    bind_rows()

dat_csv  <- paste0("data/lit_review/", str_subset(fls, ".csv")) %>% # head() %>% 
    map(read.csv) %>% 
    map(janitor::clean_names) %>% 
    map(function(x) x %>% select(abstract, title, year, doi, cited_by) %>% as_tibble()) %>% 
    map(function(x) x %>% mutate(across(.cols = everything(), .fns = as.character))) %>% 
    bind_rows()

dat <- bind_rows(dat_csv, dat_xls)
rm(dat_csv, dat_xls)

dat %>% 
  select(title, year, cited_by, doi, abstract) %>% 
  arrange(desc(cited_by))
```

@@

I have done the topic modelling on the data Emmy downloaded. Here some prelim results:

```{r}
## This does not compile to HTML due to memory, I compute stuff offline and add figs manually
# load("data/210903_topic_models_gibbs.RData")
# dtm

@@

Now we have a document term matrix with 73622 documents (abstracts) and 164098 words, after cleaning for common words, stop words and numbers. I pre-computed the topic models outside this notebook because it takes some time (days) and memory, so not suitable for computing at running time. Here are some visualizations of the preliminary results:

```{r}

# k <- c(25,50,100,250,500)
# df_topic_number <- tibble(
#     topic_number = k,
#     entropy = map_dbl (topicNumber.TM, function (x)
#         mean(apply(posterior(x)$topics, 1, function (z) - sum(z * log(z)))) # maximize Entropy
#     ),
#     alpha = map_dbl(topicNumber.TM, slot, "alpha"),
#     log_lik = map_dbl(topicNumber.TM, logLik) #,  #maximize loglik
#     #perplexity = map_dbl(topicNumber.TM, perplexity) #minimize perplexity
# )
# 
# df_stats <- tibble(
#     model = names(lapply(tset.TM, logLik)),
#     loglik = as.numeric(lapply(tset.TM, logLik)), #maximize loglik
#     entropy = lapply (tset.TM, function (x) 
#         mean(apply(posterior(x)$topics,
#                    1, function (z) - sum(z * log(z))))) %>% as.numeric()#maximize ENTROPY
#     
# )
# 
# 
# perp <-  lapply(tset.TM[c(1,2,4)], perplexity)
# perp$Gibbs <- NA
# # pretty names:
# df_stats$model <- c("VEM alpha", "VEM fixed", "Gibbs", "CTM")
# 
# g1 <- df_stats %>%
#     add_column(perplexity = as.numeric(perp[c(1,2,4,3)])) %>% #minimize perplexity
#     pivot_longer(cols = 2:4, names_to = "measure", values_to = "value") %>%
#     ggplot(aes(model, value)) + 
#     geom_col() + 
#     # scale_y_continuous(labels = scales::label_scientific) +
#     facet_wrap(.~measure, scales = "free_y") +
#     labs(x = "Algorithm", y = "Value", tag = "A") +
#     theme_light(base_size = 15) + 
#     theme(axis.text.x = element_text(size = 5))
# 
# g2 <- df_topic_number %>%
#     # mutate(alpha_log = log10(alpha)) %>%
#     pivot_longer(cols = 2:last_col(), names_to = "measure", values_to = "value") %>%
#     # filter(measure != "alpha") %>%
#     ggplot(aes(as.factor(topic_number), value)) +
#     geom_col() + 
#     # scale_y_continuous(labels = scales::label_scientific) +
#     labs(x = "Number of topics", y = "Value", tag = "B") +
#     facet_wrap(.~measure, scales = "free", ncol = 4, nrow = 1) +
#     theme_light(base_size = 15)
# 
# g1/g2

![Algorithm selection](figures/topicstats.png)

The best algorithm should maximize entropy and the log-likelihood while reducing perplexity. The latter cannot be calculated for Gibbs sampling, but it seems to be the best performing algorithm. The optimal number of topics is 100, where the log-likelihood starts to decay again. For now I will be using 50 topics and Gibbs sampling for the rest of the analysis (reason: simpler visualizations).

A topic is the (posterior) probability distributions of words being likely to be representing together certain hidden structures (topics) in the text. All words in the vocabulary (N = 164098 words) have a probability of explaining a topic, here I'm only showing the top 10 words per topic:

```{r fig.width=8, fig.height=10}
# df_topics25 <- tidy(topicNumber.TM[[3]], matrix = "beta")
# 
# g_25 <- df_topics25 %>%
#     group_by(topic) %>%
#     top_n(25, beta) %>%
#     ungroup() %>%
#     arrange(topic, - beta) %>%
#     mutate(term = reorder_within(term, beta, topic)) %>%
#     ggplot(aes(term, beta)) +
#     geom_col(aes(fill = as.factor(topic)),show.legend = FALSE) +
#     coord_flip() +
#     scale_x_reordered() +
#     labs(y = "Probability of word explaining the topic", x = "Word ranking") +
#     facet_wrap(.~ topic, scales = "free_y", ncol = 10) +
#     theme_light(base_size = 5) +
#     theme(axis.text.x = element_text(size = 5))
# 
# 
# g_25
# 
# ggsave(
#     filename = "topics_100.png",
#     path = "figures/",
#     plot = g_25,
#     device = "png",
#     width = 10, height = 12, units = "in",
#     dpi = 400,
#     bg = "white"
# )
```
![Topics](figures/topics_50.png)
Now let's see how topics change over time:

```{r}
# df_documents <- tidy(topicNumber.TM[[2]], matrix = "gamma")
# 
# papers_yr <- dat %>%
#     #filter(year > 2000, year < 2021) %>%
#     group_by(year) %>%
#     tally() %>% rename(total_papers = n)
# 
# 
# g_stack <-  df_documents %>% 
#     left_join(
#         dat %>% 
#             select(abstract, title, year) %>%
#             filter(!is.na(abstract)) %>%
#             unique(),
#         by = c("document" = "title") ) %>%
#     mutate(year = as.numeric(year)) %>% 
#     filter(year > 2000, year <2021) %>%
#     group_by(topic, year) %>% 
#     summarize(gamma_yr_tp = sum(gamma), .groups = "drop") %>% 
#     group_by(year) %>%
#     mutate(total = sum(gamma_yr_tp)) %>%
#     ungroup() %>% 
#     mutate(proportion = gamma_yr_tp / total,
#            topic = as.factor(topic)) %>%
#     ggplot(aes(x = year, y = proportion, group = topic)) +
#     geom_area(aes(fill = topic, color = topic), 
#               position = "stack", show.legend = FALSE, alpha = 0.5, size = 0.25) +
#     labs(y = "Relative proportion of topics", x = "Year") +
#     theme_classic(base_size = 15) 
# 
# g_stack
# 
# ggsave(
#     filename = "topics_time.png",
#     path = "figures/",
#     plot = g_stack,
#     device = "png",
#     width = 4, height = 3, units = "in",
#     dpi = 400,
#     bg = "white"
# )
```
![Topics over time](figures/topics_time.png)
There is no particular dominance of a topic over another, each topic has received relatively similar attention in academic productivity. If you're interested in learning more about the method or applications, you can learn more [here](https://arxiv.org/abs/2012.14312) or the references therein.

@@

**This is based on the old analysis with 2000 papers, please ignore for now**

@@

```{r fig.width=8}
# library(network)
# authors <- dat %>% 
#     select(authors, doi) %>% 
#     mutate(authors = str_split(authors, pattern = ", ")) %>% 
#     unnest(cols = authors) %>% 
#     mutate(presence = 1)
# 
# mat <- authors %>% unique() %>% 
#     pivot_wider(names_from = authors, values_from = presence, values_fill = 0) %>% 
#     select(-doi) %>%
#     as.matrix()
# 
# mat <- t(mat) %*% (mat)
# 
# 
# df_authors <- as_tibble(mat, rownames = "author") 
# 
# df_authors <- df_authors %>% 
#     pivot_longer(2:last_col(), names_to = "co_author", values_to = "n_papers") %>% 
#     filter(author != co_author, n_papers > 0)
# 
# net <- df_authors %>% 
#     # filtering reduces from 232k links and >4.5k authors to 708 links shared by 233 authors
#     filter(n_papers > 1) %>% 
#     network(., directed = FALSE, matrix.type = "edgelist")
# 
# networkD3::simpleNetwork(
#     Data= df_authors %>% filter(n_papers >1) ,
#     Source = "author", Target = "co_author", 
#     zoom = TRUE, linkColour = "grey", nodeColour = "blue"
# )
# 
# ## free memory
# rm(mat, authors)

### 2021-10-27 missing_values.Rmd
updating gitignore

---
title: "Missing values"
author: "Juan Rocha"
date: "May 2021"
output:
  html_document:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
    self_contained: true

  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
---

@@

```{r fig.width=8}
g <- inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year), missing = is.na(gini_mean)) %>% 
    ggplot(aes(year, gini_mean)) +
    geom_line(aes(color = country), show.legend = FALSE, size = 0.5) 
ggplotly(g, tooltip = "country") 
```

>>>>

---
title: "Missing values"
author: "Juan Rocha"
date: "May 2021"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
  html_document:

    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
    self_contained: true
---

@@

```{r fig.width=8}
g <- inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year), missing = is.na(gini_mean)) %>% 
    ggplot(aes(year, gini_mean)) +
    geom_line(aes(color = country), show.legend = FALSE, size = 0.5) +
  theme_light(base_size = 7)
ggplotly(g, tooltip = "country") 
```
